Tracing the Hercules stream around the Galaxy

### 2010-05-19 tex/hercules.tex
groups-simulations: discussion section; acks

\documentclass[12pt,preprint]{aastex}
\usepackage{amssymb,amsmath}
%\usepackage{color,hyperref}
%\definecolor{linkcolor}{rgb}{0,0,0.25}
%\hypersetup{
%  colorlinks=true,        % false: boxed links; true: colored links
%  linkcolor=linkcolor,    % color of internal links
%  citecolor=linkcolor,    % color of links to bibliography
%  filecolor=linkcolor,    % color of file links
%  urlcolor=linkcolor      % color of external links
%}
\setlength{\emergencystretch}{2em}%No overflowing references
\newcommand{\ie}{i.e.}
\newcommand{\etal}{et al.}
\newcommand{\eg}{e.g.}
\newcommand{\eqnname}{equation}
\newcommand{\figurenames}{\figurename s}
\newcommand{\sectionname}{$\mathsection$}
\newcommand{\tvector}[1]{\boldsymbol{\vec{#1}}}
\newcommand{\vx}{\tvector{x}}
\newcommand{\vv}{\tvector{v}}
\newcommand{\Gaia}{\emph{Gaia}}
\newcommand{\apogee}{APOGEE}
\newcommand{\hipparcos}{\emph{Hipparcos}}
\newcommand{\vR}{\ensuremath{v_R}}
\newcommand{\vphi}{\ensuremath{v_{\phi}}}
\newcommand{\vo}{\ensuremath{v_0}}
\newcommand{\Ro}{\ensuremath{R_0}}
\newcommand{\ro}{\Ro}
\newcommand{\Ab}{\ensuremath{A_b}}
\newcommand{\Rb}{\ensuremath{R_b}}
\newcommand{\Omegab}{\ensuremath{\Omega_b}}
\newcommand{\fdehnen}{\ensuremath{f_{\text{Dehnen}}}}
\newcommand{\sigmaR}{\ensuremath{\sigma_R}}
\newcommand{\rE}{\ensuremath{R_e}}
\newcommand{\Lc}{\ensuremath{L_c}}
\newcommand{\Rs}{\ensuremath{R_s}}
\newcommand{\Rsigma}{\ensuremath{R_{\sigma}}}
\newcommand{\Rolr}{\ensuremath{R_{\text{OLR}}}}

@@

Since the Hercules stream is only apparent in the planar motions in
the disk \citep[\eg][]{Bovy09a}, we only consider a two-dimensional
model for the Galaxy. We use a simple power-law rotation curve to
model the axisymmetric Galactic potential
\begin{equation}
v_c = \vo(R/\Ro)^\beta \,,
\end{equation}
and for most of the simulations below we specify this model further to
a flat rotation curve ($\beta = 0$). The model for the bar is given by
the following potential
\begin{equation}
\Phi_b(R,\phi) = \Ab(t) \cos [2(\phi - \Omegab t)] \times
\left\{ \begin{array}{ll} -(\Rb/R)^3\,, & \mathrm{for}\ R \geq
  \Rb\,,\\ (R/\Rb)^3-2\,, & \mathrm{for}\ R \leq \Rb\,. \end{array}
\right.
\end{equation}
Here \Omegab\ is the pattern speed of the bar and \Rb\ is the bar
radius, which is fixed to be 80\,percent of the bar's corotation
radius. The bar is grown smoothly during a time $t_1$, always set to
half of the total integration time in what follows, following the
prescription
\begin{equation}
\Ab(t) = \alpha \, \frac{\vo^2}{3}\,\left(\frac{\Ro}{\Rb}\right)^3
\times \left\{ \begin{array}{lll}
\frac{3}{16}\xi^5 - \frac{5}{8} \xi^3 + \frac{15}{16} \xi + \frac{1}{2}\,, & \xi \equiv 2\frac{t}{t_1}-1\,, & t \leq t_1\,,\\
1\,, & & t \geq t_1\,.\end{array} \right. 
\end{equation}

For the distribution function of the stellar disk before bar
formation, we use a Dehnen distribution function \citep{dehnen99b}
given by
\begin{equation}
\fdehnen(E,L) \propto \frac{\Sigma(\rE)}{\sigmaR^2(\rE)} \, \exp\left[ \frac{\Omega(\rE)\left[L-\Lc(E)\right]}{\sigmaR^2(\rE)}\right]\,,
\end{equation}
where \rE, \Lc, and $\Omega(\rE)$ are the radius, the angular
momentum, and the angular frequency, respectively, of the circular
orbit with energy $E$. Using the procedure given in \sectionname~3.2
of \citet{dehnen99b}, we choose the $\Sigma(R)$ and $\sigmaR$
functions such that they reproduce a disk with exponential surface
density and velocity dispersion profiles 
\begin{equation}
\Sigma(R) = \Sigma_0 \exp\left(-R/\Rs\right)\,, \qquad 
\sigmaR(R) = \sigma_0 \exp\left(-R/\Rsigma\right)\,,
\end{equation}
to an accuracy of a fraction of a percent at all radii. Our fiducial
model for the distribution function has \Rs = \Ro /3 and \Rsigma =
3\Rs, and $\sigma_0$ such that $\sigmaR(\Ro) = 0.2 \vo$.

@@

@@

\section{The line-of-sight velocity distribution}

@@

@@

SDSS-III Project Description\footnote{Available at
  \url{http://sdss3.org/collaboration/description.pdf}~.}

\acknowledgements David Hogg, Michael Blanton.

@@

@@

@@

\clearpage
\begin{figure}
\includegraphics[width=\textwidth]{rphi2d.ps}
\caption{Two-dimensional velocity distributions along the Solar
  circle. The axes of each subpanel are the radial and tangential
  velocities in Galactocentric cylindrical coordinates for the same
  ranges as in \figurename~\ref{fig:obs}. The velocity distributions
  on the other side of the Galaxy are obtained by shifting the
  azimuths by 180$^{\circ}$. }\label{fig:rphi2d}
\end{figure}

>>>>

\documentclass[12pt,preprint]{aastex}
\usepackage{amssymb,amsmath}
%\usepackage{color,hyperref}
%\definecolor{linkcolor}{rgb}{0,0,0.25}
%\hypersetup{
%  colorlinks=true,        % false: boxed links; true: colored links
%  linkcolor=linkcolor,    % color of internal links
%  citecolor=linkcolor,    % color of links to bibliography
%  filecolor=linkcolor,    % color of file links
%  urlcolor=linkcolor      % color of external links
%}
\setlength{\emergencystretch}{2em}%No overflowing references
\newcommand{\ie}{i.e.}
\newcommand{\etal}{et al.}
\newcommand{\eg}{e.g.}
\newcommand{\eqnname}{equation}
\newcommand{\figurenames}{\figurename s}
\newcommand{\sectionname}{$\mathsection$}
\newcommand{\tvector}[1]{\boldsymbol{\vec{#1}}}
\newcommand{\vx}{\tvector{x}}
\newcommand{\vv}{\tvector{v}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\Gaia}{\emph{Gaia}}
\newcommand{\apogee}{APOGEE}
\newcommand{\hipparcos}{\emph{Hipparcos}}
\newcommand{\vR}{\ensuremath{v_R}}
\newcommand{\vphi}{\ensuremath{v_{\phi}}}
\newcommand{\vo}{\ensuremath{v_0}}
\newcommand{\Ro}{\ensuremath{R_0}}
\newcommand{\ro}{\Ro}
\newcommand{\Ab}{\ensuremath{A_b}}
\newcommand{\Rb}{\ensuremath{R_b}}
\newcommand{\Omegab}{\ensuremath{\Omega_b}}
\newcommand{\fdehnen}{\ensuremath{f_{\text{Dehnen}}}}
\newcommand{\sigmaR}{\ensuremath{\sigma_R}}
\newcommand{\rE}{\ensuremath{R_e}}
\newcommand{\Lc}{\ensuremath{L_c}}
\newcommand{\Rs}{\ensuremath{R_s}}
\newcommand{\Rsigma}{\ensuremath{R_{\sigma}}}
\newcommand{\Rolr}{\ensuremath{R_{\text{OLR}}}}

@@

Since the Hercules stream is only apparent in the planar motions in
the disk \citep[\eg,][]{Bovy09a}, we only consider a two-dimensional
model for the Galaxy. We use a simple power-law rotation curve to
model the axisymmetric Galactic potential
\begin{equation}
v_c = \vo(R/\Ro)^\beta \,,
\end{equation}
and for most of the simulations below we specify this model further to
a flat rotation curve ($\beta = 0$). The model for the bar is given by
the following potential
\begin{equation}
\Phi_b(R,\phi) = \Ab(t) \cos [2(\phi - \Omegab t)] \times
\left\{ \begin{array}{ll} -(\Rb/R)^3\,, & \mathrm{for}\ R \geq
  \Rb\,,\\ (R/\Rb)^3-2\,, & \mathrm{for}\ R \leq \Rb\,. \end{array}
\right.
\end{equation}
Here \Omegab\ is the pattern speed of the bar and \Rb\ is the bar
radius, which is fixed to be 80\,percent of the bar's corotation
radius. The bar is grown smoothly during a time $t_1$, always set to
half of the total integration time in what follows, following the
prescription
\begin{equation}
\Ab(t) = \alpha \, \frac{\vo^2}{3}\,\left(\frac{\Ro}{\Rb}\right)^3
\times \left\{ \begin{array}{lll}
\frac{3}{16}\xi^5 - \frac{5}{8} \xi^3 + \frac{15}{16} \xi + \frac{1}{2}\,, & \xi \equiv 2\frac{t}{t_1}-1\,, & t \leq t_1\,,\\
1\,, & & t \geq t_1\,.\end{array} \right. 
\end{equation}

For the distribution function of the stellar disk before bar
formation, we use a Dehnen distribution function \citep{dehnen99b}
given by
\begin{equation}\label{eq:fdehnen}
\fdehnen(E,L) \propto \frac{\Sigma(\rE)}{\sigmaR^2(\rE)} \, \exp\left[ \frac{\Omega(\rE)\left[L-\Lc(E)\right]}{\sigmaR^2(\rE)}\right]\,,
\end{equation}
where \rE, \Lc, and $\Omega(\rE)$ are the radius, the angular
momentum, and the angular frequency, respectively, of the circular
orbit with energy $E$. Using the procedure given in \sectionname~3.2
of \citet{dehnen99b}, we choose the $\Sigma(R)$ and $\sigmaR$
functions such that they reproduce a disk with exponential surface
density and velocity dispersion profiles 
\begin{equation}
\Sigma(R) = \Sigma_0 \exp\left(-R/\Rs\right)\,, \qquad 
\sigmaR(R) = \sigma_0 \exp\left(-R/\Rsigma\right)\,,
\end{equation}
to an accuracy of a fraction of a percent at all radii. Our fiducial
model for the distribution function has \Rs = \Ro /3 and \Rsigma =
3\Rs, and $\sigma_0$ such that $\sigmaR(\Ro) = 0.2 \vo$.

@@

In what follows, all velocities are always considered to be with
respect to their local standard of rest in the absence of a bar.

@@

\section{Line-of-sight velocity distributions}

Rather than wait for a comprehensive survey of the disk of the Galaxy
that measures the full phase-space position of a large number of stars
by measuring proper motions, we can ask whether a targeted survey of
line-of-sight velocities in selected regions of the Galaxy could see a
clear signature of the Hercules feature. That this should be possible
follows from the following argument. The Hercules feature stands out
in the velocity distribution most in its Galactocentric tangential
velocity $\vphi$. On the Solar circle near the Sun we see almost
entirely this Galactocentric tangential velocity in the line-of-sight,
so we would expect to see at least a hint of the Hercules stream in
the line-of sight.

In \figurename~\ref{fig:rphi1d} I show the predicted line-of-sight
velocity distributions in various locations near the Solar circle all
around the Galaxy, obtained by integrating the full two-dimensional
velocity distribution over the tangential velocity. This
\figurename\ is no longer symmetric with respect to shifts of
180$^{\circ}$ degree in azimuth, since the line-of-sight direction
breaks this symmetry. The gray curve is the line-of-sight distribution
one would observe if the targeted stars were in a steady state in an
axisymmetric potential with a Dehnen distribution function
(\eqnname~[\ref{eq:fdehnen}]). Both line-of-sight velocity
distributions are normalized to have equal mass since I assume that
our knowledge of the Galaxy and the tracer population will not be
great enough that we can test predicted differences in the absolute
number of stars at a given location between the axisymmetric Galaxy
and the barred Galaxy. These predictions are also largely a function
of the particular form and parameters of the distribution function
that we chose the represent the stellar disk before bar formation.

We see that the prospects to see the Hercules feature are quite good
on our side of the Galaxy. The Hercules feature is especially strong
in the line-of-sight velocity distribution for Galactic longitudes $l
\geq 250^{\circ}$. This is because of the fortunate coincidence that
for those lines-of-sight we see velocities in the direction along
which the Hercules streams separates most from the bulk of the
velocity distribution (see \figurename~\ref{fig:rphi2d}) and that at
these azimuths the Hercules stream is at the furthest distance from
the rest of the velocity distribution. We see hardly any effect of the
bar on the line-of-sight velocity distributions on the other side of
the Galaxy ($90^{\circ} \leq \phi \leq 270^{\circ}$). This is due to
the fact that on that side we see mostly the Galactocentric radial
velocities. Even though upcoming surveys such as \apogee\ have the
ability to probe the other side of the Galaxy, for the purposes of
detecting the Hercules stream it is uninteresting.

To further analyze the best directions to target to best see the
Hercules feature in the line-of-sight velocities, I have calculated
the line-of-sight velocity distribution on a 100 by 20 grid in
Galactocentric azimuth and radius. As a measure of the difference
between the barred Galaxy model and the steady-state, non-barred
model, I have calculated the Kullback-Leibler divergence between these
predicted distributions \citep[\eg,][]{mackay}
\begin{equation}
D_{\text{KL}}(P||Q) = \int \dd x\,p(x)\,\ln \frac{p(x)}{q(x)}\,,
\end{equation}
where, in this case, $x$ is the line-of-sight velocity, $p(x)$ is the
predicted distribution for the bar-origin model of the Hercules
stream, and $q(x)$ is the axisymmetric prediction. The
Kullback-Leibler divergence is only one among many possible choices to
quantify the difference between two probability distributions. It can
be interpreted as the expected weight of evidence for the prediction
by the barred Galaxy model over the fully axisymmetric prediction that
can be expected from each sampled star. As such, locations with the
maximal Kullback-Leibler divergence are those in which the Hercules
feature stands out most prominent in the line-of-sight.

The result of this analysis is shown in
\figurename~\ref{fig:detect}. I present results both using the full
line-of-sight velocity distributions and only considering the part of
the distribution at $|v_{\text{los}}| \geq 0.15 \vo$. As is clear from
\figurename~\ref{fig:obs}, the local velocity distribution contains
many more features than the Hercules stream. These features are mostly
contained to the main mode of the velocity distribution $|v| \lesssim
0.15 \vo$. Observations indicate that these features are also the
result of dynamical interactions \citep[\eg,][]{Bovy10a,sellwood10a},
albeit with spiral arms rather than the bar. Therefore, we can expect
these or similar features to be present in the velocity distribution
at other locations in the Galaxy as well, and thus, if the effect of
the Hercules stream occurs close to the center of the bulk of the
velocity distribution it might be hard to disentangle from the other
dynamical effects at play. However, these additional dynamical effects
will be much less important for velocities $\gtrsim 0.15
\vo$. \figurename~\ref{fig:detect} shows that there are many
locations, mostly at $l \gtrsim 250^{\circ}$, where the Hercules
stream is clearly offset from the bulk of the velocity distribution
and should be unambiguously detectable.

Finally, in \figurename~\ref{fig:1dvar} I show the influence of
various nuisance and dynamical parameters on the predicted
line-of-sight velocity distribution for the most promising
line-of-sight and distance from the Sun that is more than 3 kpc
away. We see that the effect of distance uncertainties is
negligible. The exact shape and parameters of the stellar distribution
function also only minimally affect the predicted line-of-sight
velocity distribution. Changing the scalelength within its current
uncertainty changes the prediction only in the amplitude of the
effect: The Hercules stream is more prominent for shorter scalelengths
since then there are relatively more stars in the inner Galaxy that
are brought to the Solar circle by the interaction with the bar. Using
a Shu distribution function \citep{shu69a} instead of the Dehnen
distribution function also gives very similar predictions. Changing
the axisymmetric potential by changing the shape of the rotation curve
shifts the peak as expected \citep[see][]{dehnen00a}, but the Hercules
feature nevertheless remains detectable.

@@

The astrometric \Gaia\ mission will measure the parallaxes and proper
motions of up to one billion stars, most of which will be disk
stars. With parallax and proper motion accuracies down to 10
microarcseconds, \Gaia\ will be able to probe the kinematics of the
disk out to several kpc in all directions
\citep{bailerjones08a}. Crucially, it will provide large samples of
stars in many of the promising regions in
\figurename~\ref{fig:rphi2d}, which will be able to verify the
predictions laid out in this paper and will allow the dynamical
properties of the bar to be tightly constrained, something that will
be hard for \Gaia\ to do using direct observations of the bulge due to
the large extinction toward the Galactic center.

@@

Since the extinction in the optical bandpasses is rather large in the
directions identified as promising in \figurename~\ref{fig:detect}
\citep[$A_V \approx 8$ mag toward $l \approx
  270^{\circ}$;][]{marshall06a}, spectroscopic surveys in the
near-infrared such as \apogee\ are the best bet for seeing the
Hercules feature in the line-of-sight direction. The range of Galactic
longitudes that \apogee\ can target along the disk is $-5^{\circ} \leq
l \leq 250^{\circ}$ (SDSS-III Project Description\footnote{Available
  at \url{http://sdss3.org/collaboration/description.pdf}~.}). From
\figurename~\ref{fig:detect} this limit seems to be right on the edge
of where one might hope to detect the Hercules feature. Since
\apogee\ is expected to start taking data in 2011 and to finish
operations before the first \Gaia\ data release, \apogee\ could trace
the Hercules stream around the Galaxy \Gaia. Since \apogee\ will also
obtain detailed chemical abundances for its target stars, this would
also allow for a more detailed analysis of the abundances of stars in
the Hercules stream \citep[following, \eg,][]{Bensby07a,Bovy10a},
which could shed more light on radial mixing in the Galactic disk.

\acknowledgements It is a pleasure to thank Mike Blanton and David
Hogg for helpful discussions and assistance.  Financial support for
this project was provided by the National Aeronautics and Space
Administration (grant NNX08AJ48G) and the National Science Foundation
(grant AST-0908357).

@@

\bibitem[Mackay(2003)]{mackay}
  Mackay,~D.~J.~C., 2003,
  Information theory, inference, and learning algorithms (Cambridge University Press)

\bibitem[Marshall \etal(2006)]{marshall06a}
  Marshall,~D.~J.~, \etal, 2006,
  \aap, 453, 635

@@

\bibitem[Sellwood(2010)]{sellwood10a}
  Sellwood,~J.~A., 2010,
  \mnras, submitted, 
  arXiv:1001.5197v1 [astro-ph.GA]

@@

\clearpage
\begin{figure}
\includegraphics[width=\textwidth]{rphi2d.ps}
\caption{Two-dimensional velocity distributions along the Solar
  circle. The axes of each subpanel are the radial and tangential
  velocities in Galactocentric cylindrical coordinates for the same
  ranges as in \figurename~\ref{fig:obs}. The velocity distributions
  on the other side of the Galaxy are obtained by shifting the
  azimuths by 180$^{\circ}$.}\label{fig:rphi2d}
\end{figure}
